\section{Bitmaps}

TinyVicky allows for two full screen bitmaps to be displayed at once. These bitmaps are either $320 \times 200$ or $320 \times 240$, depending on the value of the CLK\_70 bit of the master control register. A bitmap's pixel data contains either 64,000 bytes, or 76,800 bytes of data. In both cases, the pixel data is arranged from left to right and top to bottom. The first 320 bytes are the pixels of the first line (with the first pixel being the left-most). The second 320 bytes are the second line, and so on. Additionally, the bitmaps can independently use any of the four graphics CLUTs to specify the colors for those indexes. TinyVicky provides registers for each bitmap set the CLUT and the address of the bitmap:

\begin{table}[h]
    \begin{center}
        \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|} \hline
            Address & Bitmap & 7 & 6 & 5 & 4 & 3 & 2 & 1 & 0 \\ \hline\hline
            \verb+0xD100+ & \multirow{4}{*}{0} & \multicolumn{5}{|c|}{---} & \multicolumn{2}{|c|}{CLUT} & ENABLE \\\cline{1-1}\cline{3-10}
            \verb+0xD101+ & & AD7 & AD6 & AD5 & AD4 & AD3 & AD2 & AD1 & AD0 \\\cline{1-1}\cline{3-10}
            \verb+0xD102+ & & AD15 & AD14 & AD13 & AD12 & AD11 & AD10 & AD9 & AD8 \\\cline{1-1}\cline{3-10}
            \verb+0xD103+ & & \multicolumn{6}{|c|}{---} & AD17 & AD16 \\ \hline
            \verb+0xD108+ & \multirow{4}{*}{1} & \multicolumn{5}{|c|}{---} & \multicolumn{2}{|c|}{CLUT} & ENABLE \\\cline{1-1}\cline{3-10}
            \verb+0xD109+ & & AD7 & AD6 & AD5 & AD4 & AD3 & AD2 & AD1 & AD0 \\\cline{1-1}\cline{3-10}
            \verb+0xD10A+ & & AD15 & AD14 & AD13 & AD12 & AD11 & AD10 & AD9 & AD8 \\\cline{1-1}\cline{3-10}
            \verb+0xD10B+ & & \multicolumn{6}{|c|}{---} & AD17 & AD16 \\ \hline
        \end{tabular}
    \end{center}
    \caption{Bitmap Registers}
    \label{tab:bm_registers}
\end{table}

\begin{description}
    \item[ENABLE] if set and both graphics and bitmaps are enabled in the Vicky Master Control Register (see table~\ref{tab:vky_master_ctrl_reg}), then this bitmap will be displayed.

    \item[CLUT] sets the graphics color lookup table to be used for this bitmap

    \item[AD] give the address of the first byte of the pixel data within the 256 KB system RAM. Note that this address is relative to the system bus of 20 bits and is not based on the CPU's addressing.
\end{description}

Setting up a bitmap is relatively straightforward. The first thing is to enable graphics and the bitmap engine in the TinyVicky Master Control Register. Then the bitmap's pixel data should be loaded into RAM somewhere. Then the address of the first byte of the bitmap pixel data is put into the address field of the bitmap control registers. The bitmap should be placed in the appropriate layer. Then, the bitmap's main control register should be set to enable it and set which graphics CLUT should be used to interpret the bitmap's color data.
